% !TEX root = licenta.tex
\documentclass[11pt]{report} 
\usepackage{geometry} % Load geometry early to prevent clash 

% Language may be romanian or english (default is english)
% Type may be bachelor or master (default is bachelor)
\usepackage[language=english, type=bachelor]{style}

% --- OVERRIDING STYLE.STY HARDCODES ---
\geometry{top=2cm, bottom=2cm, left=2.5cm, right=2cm}
\linespread{1.15}
% --------------------------------------

% Controlling the appearance of your headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\chead{}
\renewcommand{\headrulewidth}{0.2pt}
\renewcommand{\footrulewidth}{0.2pt}

% ==========================================
% CUSTOM THESIS DEPENDENCIES INCORPORATED
% ==========================================
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, calc, backgrounds}

% Define brand colors for code block aesthetics
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{codeblue}{rgb}{0,0.2,0.6}
\definecolor{backcolour}{rgb}{0.96,0.96,0.95}

% Define the structural style for \begin{lstlisting}
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen}\itshape,
    keywordstyle=\color{blue}\bfseries,
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=10pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    frame=single,
    rulecolor=\color{black!30}
}
\lstset{style=mystyle}

% Define JavaScript as a language for lstlisting (not built-in)
\lstdefinelanguage{JavaScript}{
  keywords={break, case, catch, continue, debugger, default, delete, do,
    else, finally, for, function, if, in, instanceof, new, return, switch,
    this, throw, try, typeof, var, void, while, with, const, let, class,
    export, import, extends, super, async, await, of, from, static},
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]',
  morestring=[b]",
  morestring=[b]`,
  sensitive=true
}

% Define TikZ Box Styles
\tikzstyle{abstractbox} = [rectangle, rounded corners, minimum width=3.5cm, minimum height=1cm, text centered, draw=black, fill=blue!10, font=\sffamily\footnotesize]
\tikzstyle{iobox} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3.5cm, minimum height=1cm, text centered, draw=black, fill=green!10, font=\sffamily\footnotesize]
\tikzstyle{processbox} = [rectangle, minimum width=3.5cm, minimum height=1cm, text centered, draw=black, fill=orange!10, font=\sffamily\footnotesize]
\tikzstyle{arrow} = [thick,->,>=stealth]
\tikzstyle{dashedarrow} = [thick,dashed,->,>=stealth]

\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=magenta,      
    urlcolor=blue,
}
% ==========================================

\begin{document}

\specialization{Computer Science}	
\title{Time Series Forecasting Visualizer: A Decoupled Architecture for Econometric Analytics and Deterministic Semantic Interpretation}					   
\author{Afloarei Codrin-Ionut}											
\supervisor{Petrescu Manuela}				
				
\maketitle

% Romanian Title Page
\specialization{Informatică în limba engleză}
\title{Vizualizator pentru prognoza seriilor de timp: O arhitectură decuplată pentru analiză econometrică și interpretare semantică deterministă}
\setbool{@eng}{false}
\maketitle
\setbool{@eng}{true} % Restore english mode for the rest of the document

\newpage
\thispagestyle{empty}
\mbox{}
\newpage
\pagenumbering{roman} 

\cleardoublepage
ABSTRACT (EN)
\vspace{0.5cm}	
\hrule
\vspace{0.5cm}	

Traditional econometric forecasting and structural break detection often rely on heavy, isolated computational environments, creating significant barriers to entry for non-technical end users and inducing severe latency in web-based interfaces. Furthermore, applying global statistical models across volatile macroeconomic shocks routinely yields highly inaccurate predictions. To resolve this, this thesis introduces a high-performance, decoupled web architecture designed to democratize complex time-series analytics. 

By natively porting rigorous mathematical algorithms—including Autoregressive Integrated Moving Average (ARIMA), Holt-Winters Exponential Smoothing, and Pruned Exact Linear Time (PELT) structural break detection—directly into the Node.js V8 execution engine, the system eliminates cross-language network bottlenecks and computes robust data science natively in JavaScript. When structural breaks are mathematically identified, the architecture dynamically segments the data, discarding obsolete pre-shock history to drastically improve predictive accuracy (measured via RMSE).

To translate these complex mathematical operations into accessible business intelligence, the presentation layer utilizes the React rendering lifecycle to rapidly calculate interactive Chart.js HTML5 Canvas graphics. Concurrently, the backend pipeline dynamically routes the proven mathematical accuracy gains through the Google Gemini Large Language Model. By employing strict Deterministic Prompt Engineering at a locked temperature of $0.0$, the architecture produces consistent, mathematically grounded textual insights. Ultimately, this thesis demonstrates how modern Software Engineering patterns, State Management, and asynchronous API design can converge to deliver advanced econometric analytics within a standard browser environment.

\newpage
ABSTRACT (RO)
\vspace{0.5cm}	
\hrule
\vspace{0.5cm}	

Prognoza econometrică tradițională și detecția rupturilor structurale se bazează adesea pe medii de calcul greoaie și izolate, creând bariere semnificative la intrare pentru utilizatorii finali non-tehnici și inducând o latență severă în interfețele web. Mai mult, aplicarea modelelor statistice globale peste șocuri macroeconomice volatile produce în mod curent predicții extrem de inexacte. Pentru a rezolva acest lucru, această teză introduce o arhitectură web decuplată, de înaltă performanță, concepută pentru a democratiza analiza complexă a seriilor de timp.

Prin portarea nativă a algoritmilor matematici riguroși—inclusiv Autoregressive Integrated Moving Average (ARIMA), Netezirea Exponențială Holt-Winters și detecția rupturilor structurale Pruned Exact Linear Time (PELT)—direct în motorul de execuție Node.js V8, sistemul elimină blocajele de rețea inter-limbaj și calculează știința datelor robustă nativ în JavaScript. Atunci când rupturile structurale sunt identificate matematic, arhitectura segmentează dinamic datele, eliminând istoricul învechit de dinaintea șocului pentru a îmbunătăți drastic acuratețea predictivă (măsurată prin RMSE).

Pentru a traduce aceste operațiuni matematice complexe în inteligență de afaceri accesibilă, stratul de prezentare utilizează ciclul de viață de randare React pentru a calcula rapid grafice interactive Chart.js HTML5 Canvas. Concomitent, conducta backend rutează dinamic câștigurile de acuratețe matematică dovedite prin Modelul de Limbaj Mare Google Gemini. Prin utilizarea strictă a Ingineriei Deterministe a Prompturilor la o temperatură blocată de $0.0$, arhitectura produce perspective textuale consistente, fundamentate matematic. În cele din urmă, această teză demonstrează modul în care modelele moderne de Inginerie Software, Managementul Stării și designul asincron al API-urilor pot converge pentru a livra analize econometrice avansate într-un mediu standard de browser.

\tableofcontents

\newpage
\pagenumbering{arabic}

\input{chapters/chapter1_introduction}
%\addcontentsline{toc}{chapter}{Introducere}
%\addcontentsline{toc}{chapter}{Introduction}

\input{chapters/chapter2}
\input{chapters/chapter3}
\input{chapters/chapter4}

\input{chapters/chapter6_conclusions}
%\addcontentsline{toc}{chapter}{Concluzii}
%\addcontentsline{toc}{chapter}{Conclusions}

\begin{thebibliography}{99}
\bibitem{BaiPerron1998} Bai, J., \& Perron, P. (1998). "Estimating and Testing Linear Models with Multiple Structural Changes." \textit{Econometrica}, 66(1), 47-78.
\bibitem{BoxJenkins1970} Box, G. E. P., \& Jenkins, G. M. (1970). \textit{Time Series Analysis: Forecasting and Control}. Holden-Day.
\bibitem{ClementsHendry1998} Clements, M. P., \& Hendry, D. F. (1998). \textit{Forecasting Economic Time Series}. Cambridge University Press.
\bibitem{Hamilton1994} Hamilton, J. D. (1994). \textit{Time Series Analysis}. Princeton University Press.
\bibitem{Hyndman2008} Hyndman, R. J., Koehler, A. B., Ord, J. K., \& Snyder, R. D. (2008). \textit{Forecasting with Exponential Smoothing: The State Space Approach}. Springer.
\bibitem{Killick2012} Killick, R., Fearnhead, P., \& Eckley, I. A. (2012). "Optimal detection of changepoints with a linear computational cost." \textit{Journal of the American Statistical Association}, 107(500), 1590-1598.
\bibitem{Korinek2025} Korinek, A. (2025). "AI Agents for Economic Research." \textit{Journal of Economic Literature}.
\bibitem{Lucas1976} Lucas, R. E. (1976). "Econometric Policy Evaluation: A Critique." \textit{Carnegie-Rochester Conference Series on Public Policy}, 1, 19-46.
\bibitem{Makridakis1998} Makridakis, S., Wheelwright, S. C., \& Hyndman, R. J. (1998). \textit{Forecasting: Methods and Applications} (3rd ed.). John Wiley \& Sons.
\bibitem{Pesaran2004} Pesaran, M. H., \& Timmermann, A. (2004). "Real Time Econometrics." \textit{IZA Discussion paper series}, No. 1108.
\bibitem{Schwarz1978} Schwarz, G. E. (1978). "Estimating the dimension of a model." \textit{Annals of Statistics}, 6(2), 461-464.
\bibitem{Vaswani2017} Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N., Kaiser, \L., \& Polosukhin, I. (2017). "Attention Is All You Need." \textit{Advances in Neural Information Processing Systems}, 30, 5998-6008.
\bibitem{Wold1938} Wold, H. (1938). \textit{A Study in the Analysis of Stationary Time Series}. Almqvist \& Wiksell.
\end{thebibliography}

\end{document}
